### 2020-06-17 report/sections/background/related.tex
Report: complete background

\subsubsection{Runtime Monitors in Python}
Neykova and Yoshida targeted the MPST methodology 
for Python programs in \cite{Python2017} 
and proposed to generate \textit{runtime monitors} 
from the EFSM.
These monitors expose APIs for sending and receiving messages,
which is used by the developer in their implementation. 
The runtime monitor is an abstraction between the 
developer's implementation and the actual communication channel, 
and ``executes'' the EFSM internally to ensure protocol conformance. 
When the developer sends a message (with some label and payload) 
using the API, the runtime monitor checks 
whether this send action conforms to the current EFSM state, 
and if so, 
performs the send and advances to the successor state. 
Likewise, when the developer invokes a receive, 
the runtime monitor verifies that this is permitted 
at the current EFSM state before returning the received payload.

@@

We discuss the current state-of-the-art proposals
for integrating session types in web development.

@@

This proposal is relevant to the problem we are tackling.
In particular, the final point about guaranteeing channel linearity
by construction is something we can build upon in our solution,
since linear types is a feature left to be desired in
``mainstream'' programming languages.
We also observe the challenges of applying session types 
to the front-end environment, as shown by the careful choice 
made in \cite{PureScript2019} 
to use the \textit{Concur UI} framework which 
builds UI elements sequentially to model sequential sessions.
Not doing so would require binding channel resources into 
event listeners on UI elements, which makes linearity violation possible 
(e.g. by binding a send transition to a button click event,
the button may still be active after the transition is complete;
clicking on it again will reuse the channel resource).

As motivated in \cref{section:intro}, \dots

\subsubsection{MVU using Links}
\dots

>>>>

\subsubsection{Runtime Monitors in Python}
Neykova targeted the MPST methodology 
for Python programs in \cite{Python2017} 
and proposed to generate \textit{runtime monitors} 
from the EFSM.
These monitors expose APIs for sending and receiving messages,
which is used by the developer in their implementation. 
The runtime monitor is an abstraction between the 
developer's implementation and the actual communication channel, 
and ``executes'' the EFSM internally to ensure protocol conformance. 
When the developer sends a message (with some label and payload) 
using the API, the runtime monitor checks 
whether this send action conforms to the current EFSM state, 
and if so, 
performs the send and advances to the successor state. 
Likewise, when the developer invokes a receive, 
the runtime monitor verifies that this is permitted 
at the current EFSM state before returning the received payload.

@@

We discuss the current state-of-the-art proposals
for session-typed web development.

@@

This proposal is relevant to the problem we are tackling.
In particular, the final point about guaranteeing channel linearity
by construction is something we can build upon in our solution,
since linear types is a feature left to be desired in
``mainstream'' programming languages.
We also observe the challenges of session-typed GUI
programming, as shown by the careful choice 
made in \cite{PureScript2019} 
to use the \textit{Concur UI} framework \cite{ConcurUI} which 
builds UI elements sequentially to model sequential sessions.
Not doing so would make channel linearity violation possible 
(e.g. by binding a send transition to a button click event,
the button may still be active after the transition is complete;
clicking on it again will reuse the channel resource).
Whilst \cite{PureScript2019} supports multiparty session types,
the communication protocol must describe all interactions
to go through a centralised server role to respect the
network topology enforced by WebSocket transport.

As motivated in \cref{section:intro}, 
the functional paradigm and sequential UI framework
are necessary features for the proposal to statically 
provide communication safety guarantees, but 
this type of workflow is not compatible with 
modern web programming practices and compromises on usability.

\subsubsection{Session-Typed GUI Programming using Links}
Fowler presented the first formal integration
of session typing and GUI programming in \cite{MVU2020}.
His work formalised the \textit{Model-View-Update} (MVU)
architectural pattern for GUI development.
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
user interface, and an
\textit{update function} handling
\textit{messages} produced by the
rendered model to produce a new model.
Fowler introduces the concept of model types
to express type dependencies between these components:
a \emph{model type} uniquely defines a \textit{view} function,
set of \textit{messages} and \textit{update function} --
rather than producing a new model, the update function 
defines valid transitions to other model types.
This approach guarantees channel linearity for GUI programming:
the update function transitions to a successor model, which only
renders an interface with the set of messages unique to that model,
so channel actions from the predecessor model are not rendered,
which makes reuse impossible.
The core MVU calculus is also extended with \textit{commands}
and \textit{linearity} required to express communication
actions in session type theory.

\cite{MVU2020} is implemented using
the \textit{Links} web programming language
\cite{LINKS}. Links is a strict, typed, functional language
that can be used to write full-stack web applications;
and similar to PureScript, it is compiled to JavaScript.
As it has a linear type system, session types can be
implemented using native language features, since the
language can monitor linear channel usage.
Links also supports error handling for session types,
as introduced in \cite{Exceptional}; 
given side-effects must be made explicit
in functional programming, \cite{Exceptional} introduces
a \textit{session cancellation} operator to ensure
that channel resources are closed after handling
side-effects, such as errors.

This work succeeds in 
retaining the declarative approach for writing
web interfaces, as the view function incorporates standard
\textit{Hyper-Text Mark-up Language} (HTML).
The novel approach of guaranteeing channel linearity
in GUI programming through the model type abstraction
is something we can build upon. 
However, Links does not support multiparty session types,
and similar to \cite{PureScript2019},
it can only express server-centric communication structures
over WebSocket transport.

### 2020-06-17 report/sections/server/react/approach.tex
Report: complete background

\section{Approach}
We motivate our approach from \cite{MVU2020} by extending
their work on \textit{multiple model types} motivated by the
\textit{Model-View-Update} architecture (MVU).
An MVU application features a \textit{model} encapsulating application
state, a \textit{view function} rendering the state on the 
Document Object Model (DOM), and an
\textit{update function} handling \textit{messages} produced by the
rendered model to produce a new model.
The concept of model types express type dependencies between these
components: a \emph{model type} uniquely defines a \textit{view function},
set of \textit{messages} and \textit{update function} -- rather than
producing a new model, the update function defines valid transitions to
other model types.

>>>>

\section{Approach}
We motivate our approach from \cite{MVU2020} by extending
their work on \textit{multiple model types} motivated by the
\textit{Model-View-Update} architecture (MVU),
introduced in \cref{section:bgrelated}.
The concept of model types express type dependencies between these
components: a \emph{model type} uniquely defines a \textit{view function},
set of \textit{messages} and \textit{update function} -- rather than
producing a new model, the update function defines valid transitions to
other model types.
